---
title: "openxlsx2 charts manual"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{openxlsx2 charts manual}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---


```{r setup, include = FALSE}
library(openxlsx2)
options(rmarkdown.html_vignette.check_title = FALSE)
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

The following manual will present various ways to add plots and charts to `openxlsx2` worksheets and even chartsheets. This assumes that you have basic knowledge how to handle `openxlsx2` and are familiar with either the default `R` `graphics` functions like `plot()` or `barplot()` and `grDevices`,  or with the packages `{ggplot2}`, `{rvg}` or `{mschart}`. There are plenty of other manuals that cover using these better than we could ever tell you to.

#

```{r package}
library(openxlsx2) # openxlsx2 >= 0.4 for mschart and rvg support

## create a workbook
wb <- wb_workbook()
```

## Add plot to workbook

You can include any image in PNG or JPEG format. Simply open a device and save the output and pass it to the worksheet with `wb_add_image()`.

```{r plot}
myplot <- tempfile(fileext = ".jpg")
jpeg(myplot)
print(plot(AirPassengers))
dev.off()

# Add basic plots to the workbook
wb$add_worksheet("add_image")$add_image(file = myplot)
```

## Add `{ggplot2}` plot to workbook

You can include `{ggplot2}` plots similar to how you would include them with `openxlsx`. Call the plot first and afterwards use `wb_add_plot()`.

```{r ggplot}
if (requireNamespace("ggplot2")) {

library(ggplot2)

print(ggplot(mtcars, aes(x = mpg, fill = as.factor(gear))) +
  ggtitle("Distribution of Gas Mileage") +
  geom_density(alpha = I(.5)))

# Add ggplot to the workbook
wb$add_worksheet("add_plot")$
  add_plot(width = 5, height = 3.5, fileType = "png", units = "in")

}
```


## Add plot via `{rvg}`

If you want vector graphics that can be modified in spreadsheet software the `dml_xlsx()` device comes in handy. You can pass the output via `wb_add_drawing()`.

```{r rvg}
if (requireNamespace("ggplot2") && requireNamespace("rvg")) {

library(rvg)

## create rvg example
tmp <- tempfile(fileext = ".xml")
dml_xlsx(file =  tmp, fonts = list(sans = "Bradley Hand"))
print(ggplot(data = iris,
       mapping = aes(x = Sepal.Length, y = Petal.Width)) +
  geom_point() + labs(title = "With font Bradley Hand") +
  theme_minimal(base_family = "sans", base_size = 18))
dev.off()

# Add rvg to the workbook
wb$add_worksheet("add_drawing")$
  add_drawing(xml = tmp)$
  add_drawing(xml = tmp, dims = NULL)

}
```

## Add `{mschart}` plots

If you want native open xml charts, have a look at `{mschart}`. Create one of the chart files and pass it to the workbook with `wb_add_mschart()`. There are two options possible. 
1. Either the default `{mschart}` output identical to the one in `{officer}`. Passing a data object and let `{mschart}` prepare the data. In this case `wb_add_mschart()` will add a new data region.
2. Passing a `wb_data()` object to `{mschart}`. This object contains references to the data on the worksheet and allows using data "as is".

```{r mschart}
if (requireNamespace("mschart")) {

library(mschart) # mschart >= 0.4 for openxlsx2 support

## create chart from mschart object (this creates new input data)
mylc <- ms_linechart(
  data = browser_ts,
  x = "date",
  y = "freq",
  group = "browser"
)

wb$add_worksheet("add_mschart")$add_mschart(dims = "A10:G25", graph = mylc)


## create chart referencing worksheet cells as input
# write data starting at B2
wb$add_worksheet("add_mschart - wb_data")$
  add_data(x = mtcars, dims = "B2")$
  add_data(x = data.frame(name = rownames(mtcars)), dims = "A2")

# create wb_data object this will tell this mschart
# from this PR to create a file corresponding to openxlsx2
dat <- wb_data(wb, dims = "A2:G10")

# create a few mscharts
scatter_plot <- ms_scatterchart(
  data = dat,
  x = "mpg",
  y = c("disp", "hp")
)

bar_plot <- ms_barchart(
  data = dat,
  x = "name",
  y = c("disp", "hp")
)

area_plot <- ms_areachart(
  data = dat,
  x = "name",
  y = c("disp", "hp")
)

line_plot <- ms_linechart(
  data = dat,
  x = "name",
  y = c("disp", "hp"),
  labels = c("disp", "hp")
)

# add the charts to the data
wb <- wb %>%
  wb_add_mschart(dims = "F4:L20", graph = scatter_plot) %>%
  wb_add_mschart(dims = "F21:L37", graph = bar_plot) %>%
  wb_add_mschart(dims = "M4:S20", graph = area_plot) %>%
  wb_add_mschart(dims = "M21:S37", graph = line_plot)

# add chartsheet
wb <- wb %>%
  wb_add_chartsheet() %>%
  wb_add_mschart(graph = scatter_plot)

}
```

